Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-06-23 doc/thesis/intro.tex
Most thesis changes done

%Although phylodynamics is quite new, these phenomena have been studied in
%evolutionary biology for some time. Viral phylogenies are a specific version of
%a more general class of trees called \defn{gene trees}, which represent the
%evolutionary history of a section of genetic material. Transmission trees, on
%the other hand, are highly analogous to \defn{species trees}, whose tips are
%species and internal nodes are common ancestors. This analogy derives from the
%functional similarity between transmission and allopatric speciation. Hence,
%the potential discordance between transmission trees and viral phylogenies is
%the similar to that between gene and species trees, which is called
%\defn{incomplete lineage sorting}.

%What we have just discussed is a two-step procedure for estimating the
%transmission tree. First, a viral phylogeny is constructed from genetic
%sequence data, and then it is rooted and time-scaled into a transmission tree.
%This approach is straightforward, frequently used, and has the advantage of
%leveraging tried-and-true tools for phylogenetic inference. However, it also
%has drawbacks, perhaps the most obvious being the multiplication of errors
%produced by the separate steps. One commonly used alternative method is to
%directly estimate a time-scaled phylogeny by simultaneously inferring the tree
%topology, its root and branch lengths, and the parameters of a \defn{molecular
%clock} model. A molecular clock is a hypothesis about the evolutionary rates
%along the branches of the tree, such as that they are all equal (a \defn{strict
%clock}) or that they are \gls{iid} from a common distribution (a \defn{relaxed
%clock}). This inference is usually done in a Bayesian framework using
%\gls{MCMC}, so that prior information (including the tip dates) can be included
%in the analysis, and the so-called nuisance parameters of the molecular clock
%model can be marginalized out. Software packages for performing these analyses
%include BEAST~\autocite{bouckaert2014beast} and
%MrBayes~\autocite{ronquist2012mrbayes}.

%Other methods are tailor-made for inferring transmission trees.
%\textcite{didelot2014bayesian} develop a Bayesian approach which allows
%transmissions to occur anywhere along the
%branches of a transmission tree, rather than being constrained to the branching
%points in the viral phylogeny. The method requires sampling of every infected
%individual, although the authors indicate that it could be extended to relax
%this assumption. \textcite{cottam2008integrating} describe a likelihood-based
%method which enumerates all transmission trees consistent with an established
%phylogeny, assigning each a likelihood based on other epidemiological data.
%This approach is novel in its integration of data from multiple sources,
%however because it enumerates a large portion of the tree space, it is unlikely
%to scale to larger epidemics. \textcite{ypma2012unravelling} develop a joint
%likelihood function integrating temporal, geographic, and genetic observations,
%and use Bayesian \gls{MCMC} to estimate both the tree and the parameters of the
%likelihood function. Their approach can handle missing data and produces high
%resolution transmission trees when multiple types of data are available. A
%different approach is undertaken by \textcite{jombart2011reconstructing}, who
%describe a method to build transmission trees directly from sequence data,
%contingent on the common ancestors also being sampled. This makes the method
%attractive for slow-evolving pathogens, but less practical for viral outbreaks
%where samples from common ancestors are unlikely to be available.

@@

\textcite{poon2013mapping} developed an alternative pairwise measure which
applies the concept of a \defn{kernel function} to phylogenies. Kernel
functions, originally developed for \glspl{SVM}~\autocite{burges1998tutorial}, 
compare objects in a space $\X$ by mapping them into a feature space $\F$ of
high or infinite dimension via a function $\varphi$. The similarity between the
objects is defined as
\[  
  K(x, x') = \langle \varphi(x), \varphi(x')\rangle,
\]
that is, the inner product of the objects' representations in the feature
space. Computing $\varphi(x)$ may be computationally prohibitive due to the
dimension of $\F$. The utility of a kernel function $K$ is that it is
constructed in such a way that it can compute the inner product without
explicitly computing $\varphi(x)$. The kernel function developed in
\autocite{poon2013mapping} will henceforth be referred to as the \defn{tree
kernel}. This kernel maps trees into the space of all possible possible
\defn{subset trees}, which are subtrees that do not necessarily extend all the
way to the tips. The subset-tree kernel was originally developed for comparing
parse trees in natural language processing~\autocite{collins2002new} and did
not incorporate branch length information. The version developed by
\textcite{poon2013mapping} includes a radial basis function to compare the
differences in branch lengths, thus incorporating both the trees' topologies
and their branch lengths in a single similarity score.

@@

The importance distribution $\eta$ constructed with \gls{SIS} is merely an
approximation to $\pi$, and may be a fairly poor one in practice depending on
the application. Try as we might to keep the variances of the weights low, the
cumulative errors at each sequential step tend to push many of the weights to
very low values~\autocite{doucet2000sequential}. This results in a poor
approximation to $\pi$, since only a few particles retain high importance
weights after all $d$ sequential steps, \add{a problem known as \defn{particle
degeneracy}. To mitigate this problem, \textcite{gordon1993novel} introduced
technique they called the \defn{bootstrap filter}, which involves a resampling
of the population of particles after each sequential step in accordance with
their importance weights. A similar idea, termed \defn{particle rejuvination},
was proposed by \textcite{liu1995blind}. These approaches cause particles with
high importance weights to be replicated in the population, while particles
with low weights may be removed. After each resampling step, the importance
weights for all particles are set equal.}

@@

\add{\Acrlong{SMC}, and the \gls{SMC} sampler, were developed for sampling from
distributions which can be evaluated up to a normalizing constant. We claim,
and shall argue more thoroughly below (\cref{subsec:just}) that the posterior
distribution}
\[
  \madd{\post(\theta \mid T) \propto \lik(T \mid \theta) \prior(\theta)}
\]
\add{for a contact network model with parameters $\theta$ and an input
transmission tree $T$ does not fall in this category. Therefore, \gls{SMC}, and
other Bayesian and \gls{ML} techniques for fitting mathematical models (see
\cref{chp:prelim}), cannot be directly applied to our problem. In particular,
\gls{MCMC} and the \gls{SMC} sampler are designed for distributions $\pi$ which
can be evaluated up to a normalizing constant $Z$, that is, $\pi(x) = \gamma(x)
/ Z$. Both algorithms calculate a ratio of the form $\pi(x) / \pi(x') \propto
\gamma(x) / \gamma(x')$ for a current value $x$ and proposed updated value $x'$
- for \gls{MCMC}, this is part of the \acrlong{MH} ratio, while for the
\gls{SMC} sampler, a similar ratio is required to calculate the importance
weights. In the context of Bayesian inference, this ratio contains a likelihood
ratio, which must be calculated by computing the individual likelihoods and
dividing them. If the likeilhood is intractible, this is clearly not a viable
approach.}

>>>>

@@

\textcite{poon2013mapping} developed an alternative pairwise measure which
applies the concept of a \defn{kernel function} to phylogenies. Kernel
functions, originally developed for \glspl{SVM}~\autocite{burges1998tutorial}, 
compare objects in a space $\X$ by mapping them into a feature space $\F$ of
high or infinite dimension via a function $\varphi$. The similarity between the
objects is defined as
\[  
  K(x, x') = \langle \varphi(x), \varphi(x')\rangle,
\]
that is, the inner product of the objects' representations in the feature
space. Computing $\varphi(x)$ may be computationally prohibitive due to the
dimension of $\F$. The utility of a kernel function is that it is
constructed in such a way that it can compute the inner product without
explicitly computing $\varphi(x)$. The kernel function developed in
\autocite{poon2013mapping} will henceforth be referred to as the \defn{tree
kernel}. This kernel maps trees into the space of all possible possible
\defn{subset trees}, which are subtrees that do not necessarily extend all the
way to the tips. The subset-tree kernel was originally developed for comparing
parse trees in natural language processing~\autocite{collins2002new} and did
not incorporate branch length information. The version developed by
\textcite{poon2013mapping} includes a radial basis function to compare the
differences in branch lengths, thus incorporating both the trees' topologies
and their branch lengths in a single similarity score.

@@

The importance distribution $\eta$ constructed with \gls{SIS} is merely an
approximation to $\pi$, and may be a fairly poor one in practice depending on
the application. Try as we might to keep the variances of the weights low, the
cumulative errors at each sequential step tend to push many of the weights to
very low values~\autocite{doucet2000sequential}. This results in a poor
approximation to $\pi$, since only a few particles retain high importance
weights after all $d$ sequential steps, \add{a problem known as \defn{particle
degeneracy}. To mitigate this problem, \textcite{gordon1993novel} introduced
technique they called the \defn{bootstrap filter}, which involves a resampling
of the population of particles after each sequential step in accordance with
their importance weights. A similar idea, termed \defn{particle rejuvenation},
was proposed by \textcite{liu1995blind}. These approaches cause particles with
high importance weights to be replicated in the population, while particles
with low weights may be removed. After each resampling step, the importance
weights for all particles are set equal.}

@@

\add{\Acrlong{SMC}, and the \gls{SMC} sampler, were developed for sampling from
distributions which can be evaluated up to a normalizing constant. We claim,
and shall argue more thoroughly below (\cref{subsec:just}) that the posterior
distribution}
\[
  \madd{\post(\theta \mid T) \propto \lik(T \mid \theta) \prior(\theta)}
\]
\add{for a contact network model with parameters $\theta$ and an input
transmission tree $T$ does not fall in this category. Therefore, \gls{SMC}, and
other Bayesian and \gls{ML} techniques for fitting mathematical models (see
\cref{chp:prelim}), cannot be directly applied to our problem. In particular,
\gls{MCMC} and the \gls{SMC} sampler are designed for distributions $\pi$ which
can be evaluated up to a normalizing constant $Z$, that is, $\pi(x) = \gamma(x)
/ Z$. Both algorithms calculate a ratio of the form $\pi(x) / \pi(x') \propto
\gamma(x) / \gamma(x')$ for a current value $x$ and proposed updated value $x'$
- for \gls{MCMC}, this is part of the \acrlong{MH} ratio, while for the
\gls{SMC} sampler, a similar ratio is required to calculate the importance
weights. In the context of Bayesian inference, this ratio contains a likelihood
ratio, which must be calculated by computing the individual likelihoods and
dividing them. If the likelihood is intractable, this is clearly not a viable
approach.}
